Title Goes Here

### 2021-04-09 analysis/paper/shceof.Rmd
Cambia orden (SAM -> PSA)
Closes #36

```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
  copy() %>% 
  .[, cEOF := NULL] %>% 
  .[]

psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]

cEOF2[psa_time, on = "time"] %>% 
  .[PC != "SAM"] %>% 
  copy() %>% 
  # .[, hgt := rotate(hgt, psa_angle)] %>%
  sep_ReIm() %>% 
  ggplot(aes(value, hgt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_smooth(method = "lm") + 
  scale_y_continuous("EOF2") +
  scale_x_continuous("PSA") +
  facet_grid(part ~ PC) +
  panel_background +
  tag_facets("rc")

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered. The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2). Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon. And, in fact, suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.

@@

```{r fake_eof}
pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

set.seed(42)
fake_sam <- eof_filtered$right %>% 
  copy() %>% 
  .[, eof := sample(eof, replace = TRUE)]  %>%
  setnames("eof", "value") %>% 
  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value. To this synthetic dataset we then apply traditional EOF. By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a). The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly. Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

The three EOFs of the synthetic dataset are shown in Figure\ \@ref(fig:fake-eof) row b. The synthetic SAM can be seen in the first EOF while the PSA is explained by the second and third. However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not. The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern. Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].

This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artifact. Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM. Studies on the relationship between the SAM and the PSA pattern would be particularly difficult to interpret.

>>>>

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered. The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2). Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.

@@

```{r fake_eof}
cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
  copy() %>% 
  .[, cEOF := NULL] %>% 
  .[]

pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

set.seed(42)
fake_sam <- eof_filtered$right %>% 
  copy() %>% 
  .[, eof := sample(eof, replace = TRUE)]

cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
  sep_ReIm() %>% 
  .[, correlate(eof, hgt), by = part]

fake_sam <- fake_sam %>% 
  setnames("eof", "value") %>% 
  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index. To this synthetic dataset we then apply traditional EOF. By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a). The former one is a standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly. Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

The three EOFs of the synthetic dataset are shown in Figure\ \@ref(fig:fake-eof) row b. The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third. However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not. The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern. This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null. In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part. Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].

This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artefact. Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM. Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.

```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]

cEOF2[psa_time, on = "time"] %>% 
  .[PC != "SAM"] %>% 
  copy() %>% 
  # .[, hgt := rotate(hgt, psa_angle)] %>%
  sep_ReIm() %>% 
  ggplot(aes(value, hgt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_smooth(method = "lm") + 
  scale_y_continuous("EOF2") +
  scale_x_continuous("PSA") +
  facet_grid(part ~ PC) +
  panel_background +
  tag_facets("rc")
```
